---
title: "Peak calling for ATAC-seq bulk data"
author: "XSun"
date: "2022-08-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r, echo = FALSE}
suppressMessages(library(ArchR))
suppressMessages(library(ggplot2))
suppressMessages(library(reshape))
suppressMessages(library(gridExtra))
suppressMessages(library(genomation))
```


# Introduction

We annotated all cells as the same cluster and did peak calling on the whole data set. 

Then, we compare union set and earlier peaks called on cell-type-specific dataset.

# Peak calling

Same settings with Alan's.

```{r echo=T, eval=FALSE, message=FALSE, warning=FALSE}
projHeart<- addReproduciblePeakSet(ArchRProj = projHeart, maxPeaks = 500000,groupBy = "CLUSTER_SET", pathToMacs2 = macs2, cutOff = 0.01, verbose = T)
```

Number of cells: 26,714

Number of used cells: 16,854

203,876 peaks were found. 

# Comparing with earlier peaks

## Comparing with peaks not cell-type-specific 

The peaks for each cell type here are not cell type specific.

```{r echo=F, message=FALSE, warning=FALSE}
load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/summary_allpeaks.rdata")
DT::datatable(summary, options = list(pageLength =10))
```

The right pannel below shows the proportion of recovered and unrecovered peaks for each cell type.

The left pannel below shows the number of recovered and unrecovered peaks for each cell type.


```{r echo=F, message=FALSE, warning=FALSE, paged.print=FALSE,fig.height=10, fig.width=20}

load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/summary_allpeaks.rdata")
summary$celltype <- rownames(summary)


summary1 <- data.frame(cbind(summary$celltype, as.numeric(summary$proportion_peak_overlap), as.numeric(summary$proportion_peak_not_recover)))
colnames(summary1) <- c("celltype","T","F")
data_melt <- melt(summary1, id.vars = "celltype")
data_melt$value <- as.numeric(data_melt$value)
colnames(data_melt)[2] <- "Recovered"

p1 <- ggplot(data_melt, aes(x=celltype, y=value, fill = Recovered)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = scales::percent) +
  geom_text(aes(label = paste0(round(value,4)*100,"%")), 
            position = position_stack(vjust = 0.5), size = 5) +
  theme_bw(base_line_size =0.3) +
  theme(axis.title.x = element_text(size = 20),
        axis.text.x = element_text(size = 18, color = "black", angle = 90, vjust = 0.5, hjust= 1),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 18, color = "black"),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16),
        title = element_text(size=20)) +
  labs(x = "Cell types",
       y = "Proportion") +
  ggtitle("All peaks") + theme(plot.title = element_text(hjust = 0.5)) 

summary1 <- data.frame(cbind(summary$celltype, as.numeric(summary$num_peak_overlap), as.numeric(summary$num_peak_celltype) -  as.numeric(summary$num_peak_overlap)))
colnames(summary1) <- c("celltype","T","F")
data_melt <- melt(summary1, id.vars = "celltype")
data_melt$value <- as.numeric(data_melt$value)
colnames(data_melt)[2] <- "Recovered"

p2 <- ggplot(data_melt, aes(x=celltype, y=value, fill = Recovered)) +
  geom_bar(position = "dodge", stat = "identity") +
  geom_text(aes(label = value), 
            position = position_dodge(width = 0.9), vjust = -0.25, size = 4) +
  theme_bw(base_line_size =0.3) +
  theme(axis.title.x = element_text(size = 20),
        axis.text.x = element_text(size = 18, color = "black", angle = 90, vjust = 0.5, hjust= 1),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 18, color = "black"),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16),
        title = element_text(size=20)) +
  labs(x = "Cell types",
       y = "Number of peaks") +
  ggtitle("All peaks") + theme(plot.title = element_text(hjust = 0.5)) 


all <- grid.arrange(p1,p2, nrow = 1)
```

The plot below shows the relationship between number of peaks for each cell type and the proportion of recovered peaks in bulk data

```{r echo=F, message=FALSE, warning=FALSE, paged.print=FALSE,fig.height=5, fig.width=7}
load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/summary_allpeaks.rdata")
summary$celltype <- rownames(summary)


p <- ggplot(summary, aes(x=num_peak_celltype, y=proportion_peak_overlap, color=celltype)) +
  
  geom_point() + 
  
  theme_bw(base_line_size =0.3) +
  
  theme(axis.title.x = element_text(size = 12),
        axis.text.x = element_text(size = 10, color = "black"),
        axis.title.y = element_text(size = 12),
        axis.text.y = element_text(size = 10, color = "black"))+
  
  labs(x = "Number of peaks",
      y = "Proportion of recovered peaks",
      color = "Cell types")+
  theme(axis.title.x = element_text(size = 20),
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 18, color = "black"),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16),
        title = element_text(size=20)) 
p

```

The plot below shows relationship between peaks recovered and cell type abundance

```{r echo=F, message=FALSE, warning=FALSE, paged.print=FALSE,fig.height=10, fig.width=22}

load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/summary_allpeaks.rdata")
load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/celltypes.rdata")

summary$celltype <- rownames(summary)

for (i in 1: nrow(summary)) {
  
  summary$number_cell[i] <- sum(celltypes == summary$celltype[i])
  
}
summary$proportion_cell <- summary$number_cell/sum(summary$number_cell)


p1 <- ggplot(summary, aes(x=proportion_cell, y=proportion_peak_overlap, color=celltype)) +
  
  geom_point(size=10) + 
  
  theme_bw(base_line_size =0.3) +
  
  theme(axis.title.x = element_text(size = 24),
        axis.text.x = element_text(size = 22, color = "black"),
        axis.title.y = element_text(size = 24),
        axis.text.y = element_text(size = 22, color = "black"))+
  
  labs(x = "Proportion of cells",
       y = "Proportion of recovered peaks",
       color = "Cell types") +
  theme(axis.title.x = element_text(size = 24),
        axis.text.x = element_text(size = 22, color = "black"),
        axis.title.y = element_text(size = 24),
        axis.text.y = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 18),
        title = element_text(size=20))  +
  ggtitle("Smooth Muscle included") + theme(plot.title = element_text(hjust = 0.5)) 

#### muscle excluded

summary <- summary[!rownames(summary) == "Smooth Muscle",]
summary$proportion_cell <- summary$number_cell/sum(summary$number_cell)

p2 <- ggplot(summary, aes(x=proportion_cell, y=proportion_peak_overlap, color=celltype)) +
  
  geom_point(size=10) + 
  
  theme_bw(base_line_size =0.3) +
  
  theme(axis.title.x = element_text(size = 24),
        axis.text.x = element_text(size = 22, color = "black"),
        axis.title.y = element_text(size = 24),
        axis.text.y = element_text(size = 22, color = "black"))+
  
  labs(x = "Proportion of cells",
       y = "Proportion of recovered peaks",
       color = "Cell types") +
  theme(axis.title.x = element_text(size = 24),
        axis.text.x = element_text(size = 22, color = "black"),
        axis.title.y = element_text(size = 24),
        axis.text.y = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 18),
        title = element_text(size=20))  +
  ggtitle("Smooth Muscle excluded") + theme(plot.title = element_text(hjust = 0.5)) 


all <- grid.arrange(p1,p2, nrow = 1)
```

We also computed odd ratio according to :


$$ OR = \frac{\frac{P(OCR_{2}|OCR_{1})}{1-P(OCR_{2}|OCR_{1})}}{\frac{P(OCR_{2}|\overline{OCR_{1}})}{1-P(OCR_{2}|\overline{OCR_{1}})}} $$
where $OCR_{1}$ is the our bulk peaks and $OCR_{2}$ are peaks for each cell type (not cell type specific). We sampled 2,038,760 (203,876 *10) unoverlapping regions (width = 501 which is the same with our peaks) from whole genome (hg38, blacklist, gaps and peaks themselves were excluded) and used these regions as $\overline{OCR_{1}}$ . 


```{r echo=F, message=FALSE, warning=FALSE}
load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/random_ranges_data/OR_10times_peakitself_excluded.rdata")
DT::datatable(OR_final, options = list(pageLength =10))
```

The plots below shows the distribution of the width of the overlaps. 

```{r echo=F, message=FALSE, warning=FALSE, paged.print=FALSE,fig.height=5, fig.width=10}

setwd("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/random_ranges_data/")

gr_random <- readBed("random_peakitself_excluded_10times_shuffled_modify.bed")

load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/peak_bulk.rdata")
peak_bulk <- peak_bulk_max

load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/peak_celltype.rdata")
peak_celltype_all <- peak


#######statistics for overlaps
overlap1 <- intersect(peak_celltype_all,peak_bulk)
width <- data.frame(overlap1@ranges@width)
colnames(width) <- "width"

p1 <- ggplot(width, aes(x=width)) +
  geom_histogram(binwidth=.5, colour="black", fill="white") +
  theme(axis.title.x = element_text(size = 22),
        axis.text.x = element_text(size = 20, color = "black"),
        axis.title.y = element_text(size = 22),
        axis.text.y = element_text(size = 20, color = "black")) +
  
  theme_bw(base_line_size =0.3) +
  ggtitle("Width distribution  -- overlaps of peaks \nacross all celltype and bulk peaks") + theme(plot.title = element_text(hjust = 0.5)) 

overlap2 <- intersect(peak_celltype_all,gr_random)
width <- data.frame(overlap2@ranges@width)
colnames(width) <- "width"

p2 <- ggplot(width, aes(x=width)) +
  geom_histogram(binwidth=.5, colour="black", fill="white") +
  theme(axis.title.x = element_text(size = 22),
        axis.text.x = element_text(size = 20, color = "black"),
        axis.title.y = element_text(size = 22),
        axis.text.y = element_text(size = 20, color = "black")) +
  
  theme_bw(base_line_size =0.3) +
  ggtitle("Width distribution  -- overlaps of peaks \nacross all celltype and random peaks") + theme(plot.title = element_text(hjust = 0.5)) 

all <- grid.arrange(p1,p2, nrow = 1)


```


We also tried to sample random regions with the same size of $OCR_{1}$ and used as $\overline{OCR_{1}}$. The results are similar with earlier ones. 

```{r echo=F, message=FALSE, warning=FALSE}
load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/random_ranges_data/OR_samesize_peakitself_excluded.rdata")
DT::datatable(OR_final, options = list(pageLength =10))
```

## Comparing with cell-type-specific peaks

The peaks for each cell type here are cell type specific.

```{r echo=F, message=FALSE, warning=FALSE}
load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/summary_celltypespecific.rdata")
DT::datatable(summary, options = list(pageLength =10))
```

The right pannel below shows the proportion of recovered and unrecovered peaks for each cell type.

The left pannel below shows the number of recovered and unrecovered peaks for each cell type.


```{r echo=F, message=FALSE, warning=FALSE, paged.print=FALSE,fig.height=10, fig.width=20}

load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/summary_celltypespecific.rdata")
summary$celltype <- rownames(summary)
summary1 <- data.frame(cbind(summary$celltype, as.numeric(summary$proportion_peak_overlap), as.numeric(summary$proportion_peak_not_recover)))
colnames(summary1) <- c("celltype","T","F")
data_melt <- melt(summary1, id.vars = "celltype")
data_melt$value <- as.numeric(data_melt$value)
colnames(data_melt)[2] <- "Recovered"

p1 <- ggplot(data_melt, aes(x=celltype, y=value, fill = Recovered)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = scales::percent) +
  geom_text(aes(label = paste0(round(value,4)*100,"%")), 
            position = position_stack(vjust = 0.5), size = 5) +
  theme_bw(base_line_size =0.3) +
  theme(axis.title.x = element_text(size = 20),
        axis.text.x = element_text(size = 18, color = "black", angle = 90, vjust = 0.5, hjust= 1),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 18, color = "black"),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16),
        title = element_text(size=20)) +
  labs(x = "Cell types",
       y = "Proportion") +
  ggtitle("Cell type specific peaks") + theme(plot.title = element_text(hjust = 0.5)) 



load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/summary_celltypespecific.rdata")
summary$celltype <- rownames(summary)
summary1 <- data.frame(cbind(summary$celltype, as.numeric(summary$num_peak_overlap), as.numeric(summary$num_peak_celltype) -  as.numeric(summary$num_peak_overlap)))
colnames(summary1) <- c("celltype","T","F")
data_melt <- melt(summary1, id.vars = "celltype")
data_melt$value <- as.numeric(data_melt$value)
colnames(data_melt)[2] <- "Recovered"

p2 <- ggplot(data_melt, aes(x=celltype, y=value, fill = Recovered)) +
  geom_bar(position = "dodge", stat = "identity") +
  geom_text(aes(label = value), 
            position = position_dodge(width = 0.9), vjust = -0.25, size = 4) +
  theme_bw(base_line_size =0.3) +
  theme(axis.title.x = element_text(size = 20),
        axis.text.x = element_text(size = 18, color = "black", angle = 90, vjust = 0.5, hjust= 1),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 18, color = "black"),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16),
        title = element_text(size=20)) +
  labs(x = "Cell types",
       y = "Number of peaks") +
  ggtitle("Cell type specific peaks") + theme(plot.title = element_text(hjust = 0.5)) 


all <- grid.arrange(p1,p2, nrow = 1)
```

The plot below shows the relationship between number of peaks for each cell type and the proportion of recovered peaks in bulk data

```{r echo=F, message=FALSE, warning=FALSE, paged.print=FALSE,fig.height=5, fig.width=7}
load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/summary_celltypespecific.rdata")
summary$celltype <- rownames(summary)

p <- ggplot(summary, aes(x=num_peak_celltype, y=proportion_peak_overlap, color=celltype)) +
  
  geom_point() + 
  
  theme_bw(base_line_size =0.3) +
  
  theme(axis.title.x = element_text(size = 12),
        axis.text.x = element_text(size = 10, color = "black"),
        axis.title.y = element_text(size = 12),
        axis.text.y = element_text(size = 10, color = "black"))+
  
  labs(x = "Number of peaks",
      y = "Proportion of recovered peaks",
      color = "Cell types") +
  theme(axis.title.x = element_text(size = 20),
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 18, color = "black"),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16),
        title = element_text(size=20)) 
p
```

The plot below shows relationship between peaks recovered and cell type abundance

```{r echo=F, message=FALSE, warning=FALSE, paged.print=FALSE,fig.height=10, fig.width=22}

load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/summary_celltypespecific.rdata")
load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/peak/celltypes.rdata")

summary$celltype <- rownames(summary)

for (i in 1: nrow(summary)) {
  
  summary$number_cell[i] <- sum(celltypes == summary$celltype[i])
  
}
summary$proportion_cell <- summary$number_cell/sum(summary$number_cell)


p1 <- ggplot(summary, aes(x=proportion_cell, y=proportion_peak_overlap, color=celltype)) +
  
  geom_point(size=10) + 
  
  theme_bw(base_line_size =0.3) +
  
    theme(axis.title.x = element_text(size = 24),
        axis.text.x = element_text(size = 22, color = "black"),
        axis.title.y = element_text(size = 24),
        axis.text.y = element_text(size = 22, color = "black"))+
  
  labs(x = "Proportion of cells",
       y = "Proportion of recovered peaks",
       color = "Cell types") +
  theme(axis.title.x = element_text(size = 24),
        axis.text.x = element_text(size = 22, color = "black"),
        axis.title.y = element_text(size = 24),
        axis.text.y = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 18),
        title = element_text(size=20))  +
  ggtitle("Smooth Muscle included") + theme(plot.title = element_text(hjust = 0.5)) 

#### muscle excluded

summary <- summary[!rownames(summary) == "Smooth Muscle",]
summary$proportion_cell <- summary$number_cell/sum(summary$number_cell)

p2 <- ggplot(summary, aes(x=proportion_cell, y=proportion_peak_overlap, color=celltype)) +
  
  geom_point(size=10) + 
  
  theme_bw(base_line_size =0.3) +
  
  theme(axis.title.x = element_text(size = 24),
        axis.text.x = element_text(size = 22, color = "black"),
        axis.title.y = element_text(size = 24),
        axis.text.y = element_text(size = 22, color = "black"))+
  
  labs(x = "Proportion of cells",
       y = "Proportion of recovered peaks",
       color = "Cell types") +
  theme(axis.title.x = element_text(size = 24),
        axis.text.x = element_text(size = 22, color = "black"),
        axis.title.y = element_text(size = 24),
        axis.text.y = element_text(size = 24, color = "black"),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 18),
        title = element_text(size=20))  +
  ggtitle("Smooth Muscle excluded") + theme(plot.title = element_text(hjust = 0.5)) 


all <- grid.arrange(p1,p2, nrow = 1)
```


Odd ratio:

```{r echo=F, message=FALSE, warning=FALSE}
load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/random_ranges_data/OR_10times_celltypespecific_peakitself_excluded.rdata")
DT::datatable(OR_final, options = list(pageLength =10))
```

The plots below shows the distribution of the width of the overlaps. 

```{r echo=F, message=FALSE, warning=FALSE, paged.print=FALSE,fig.height=5, fig.width=10}

peak_list <- readRDS("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/ArchR_Heart_Latest_Backup/PeakCalls/DA_MARKERS_FDRP_1_log2FC_1.rds")
peak_celltype_all <- peak_list@listData

allpeaks <- c(peak_list[[1]], peak_list[[2]],peak_list[[3]], peak_list[[4]],peak_list[[5]], peak_list[[6]],peak_list[[7]], peak_list[[8]])

overlap1 <- intersect(allpeaks,peak_bulk)
width <- data.frame(overlap1@ranges@width)
colnames(width) <- "width"

p1 <- ggplot(width, aes(x=width)) +
  geom_histogram(binwidth=.5, colour="black", fill="white") +
  theme(axis.title.x = element_text(size = 22),
        axis.text.x = element_text(size = 20, color = "black"),
        axis.title.y = element_text(size = 22),
        axis.text.y = element_text(size = 20, color = "black")) +
  
  theme_bw(base_line_size =0.3) +
  ggtitle("Width distribution  -- overlaps of peaks \nacross all celltype and bulk peaks") + theme(plot.title = element_text(hjust = 0.5)) 

overlap2 <- intersect(allpeaks,gr_random)
width <- data.frame(overlap2@ranges@width)
colnames(width) <- "width"

p2 <- ggplot(width, aes(x=width)) +
  geom_histogram(binwidth=.5, colour="black", fill="white") +
  theme(axis.title.x = element_text(size = 22),
        axis.text.x = element_text(size = 20, color = "black"),
        axis.title.y = element_text(size = 22),
        axis.text.y = element_text(size = 20, color = "black")) +
  
  theme_bw(base_line_size =0.3) +
  ggtitle("Width distribution  -- overlaps of peaks \nacross all celltype and random peaks") + theme(plot.title = element_text(hjust = 0.5)) 

all <- grid.arrange(p1,p2, nrow = 1)


```


We also tried to sample random regions with the same size of $OCR_{1}$ and used as $\overline{OCR_{1}}$. The results are similar with earlier ones. 

```{r echo=F, message=FALSE, warning=FALSE}
load("/project2/xinhe/xsun/heart_atlas/3.ATAC_peak/random_ranges_data/OR_samesize_celltypespecific_peakitself_excluded.rdata")
DT::datatable(OR_final, options = list(pageLength =10))
```