---
title: "Comparison with the first dataset"
author: "XSun"
date: "2022-07-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r, echo = FALSE}
suppressMessages(library(rtracklayer))
suppressMessages(library(ggplot2))
suppressMessages(library(ComplexHeatmap))
suppressMessages(library(gridExtra))
suppressMessages(library(ggvenn))
```


# Introduction

We compare the OCRs from our dataset with another [paper](https://pubmed.ncbi.nlm.nih.gov/33990324/) here.The peak calls are available [here](http://cepigenomics.org/CARE_portal/Cell_Type_Diversity.html). There are two levels of peaks- the peaks called on different clusters using MACS2 and peaks determined to be specifically accessible in a given cell type.


# Distribution of peak ranges (all peaks)

The peak ranges of our data were fixed to 501bp.

The range of the new data set:

```{r echo=F, message=FALSE, warning=FALSE, paged.print=FALSE,fig.height=10, fig.width=20}
setwd("/project2/xinhe/xsun/atlas/peakdata_ren/celltype/")
files <- list.files("/project2/xinhe/xsun/atlas/peakdata_ren/celltype/")
celltypes <- unlist(strsplit(files,split = ".narrowPeak.bed"))
extraCols_narrowPeak <- c(signalValue = "numeric", pValue = "numeric",qValue = "numeric", peak = "integer")

p <- list()
for (i in 1:length(files)){
  dat_new <- import(files[i], format = "BED",extraCols = extraCols_narrowPeak)
  dat_new <- unique(dat_new)
  
  dist <- data.frame(width(dat_new))
  colnames(dist) <- "width"
  p[[i]] <- ggplot(dist, aes(x=width)) + geom_histogram() + theme_bw(base_line_size =0.3) +
    ggtitle(celltypes[i])  +
    theme(plot.title = element_text(hjust = 0.5)) 
  
}

all <- grid.arrange(p[[1]],p[[2]],p[[3]],p[[4]],p[[5]],p[[6]],p[[7]],p[[8]], nrow = 2)
```


# Similarity between the two data sets

We first calculated the Jaccard index for each pairwise comparison between the two data sets. The results are displayed by heatmaps colored by jaccard index value.

We computed Jaccard index using:
```{r}
genomicCorr.jaccard = function(query, reference) {
  res = sum(width(intersect(query, reference))) / sum(as.numeric(width(union(query, reference))))
  return(res)
}
```

This is the heatmap for all peaks. Since the data for Lymphocyte in the new dataset is unavailable, there are just 8 columns in this plot.
```{r fig.height=5, fig.width=5, message=FALSE, warning=FALSE}
load("/project2/xinhe/xsun/atlas/jaccard_heatmap/jcd_matrix_all_uniq.rdata")
Heatmap(matrix = jcd_matrix, cluster_rows = F, 
        cluster_columns = F, show_row_names = T, row_names_side = "left",
        name = "Jaccard Index",
        col = circlize::colorRamp2(c(0, 0.3), c("white","firebrick")), ###modify the figures
        row_title = NULL,
        column_title = NULL,
        row_gap = unit(1, "mm"),
        column_gap = unit(1, "mm"),
        na_col = "white",
        use_raster = T)
```


This is the heatmap for disjoint peaks. 
```{r fig.height=5, fig.width=5, message=FALSE, warning=FALSE}
load("/project2/xinhe/xsun/atlas/jaccard_heatmap/jcd_matrix_disjoint_uniq.rdata")
Heatmap(matrix = jcd_matrix, cluster_rows = F, 
        cluster_columns = F, show_row_names = T, row_names_side = "left",
        name = "Jaccard Index",
        col = circlize::colorRamp2(c(0, 0.2), c("white","firebrick")), ###modify the figures
        row_title = NULL,
        column_title = NULL,
        row_gap = unit(1, "mm"),
        column_gap = unit(1, "mm"),
        na_col = "white",
        use_raster = T)
```


# A direct comparison between the Cardiomycyte peak sets

```{r echo=F, message=FALSE, warning=FALSE, paged.print=FALSE,fig.height=10, fig.width=10}
p<- list()

##########
ocr <- readRDS("/project2/gca/aselewa/heart_atlas_project/ArchR/ArchR_heart_latest_noAtrium/PeakCalls/DA_MARKERS_FDRP_1_log2FC_1.rds")
dat_our <- ocr[[1]]
extraCols_narrowPeak <- c(signalValue = "numeric", pValue = "numeric",
                          qValue = "numeric", peak = "integer")
dat_new <- import("/project2/xinhe/xsun/atlas/peakdata_ren/celltype/Atrial_cardiomyocyte.narrowPeak.bed", format = "BED",
                  extraCols = extraCols_narrowPeak)
dat_new2 <- unique(dat_new)

overlap <- GenomicRanges::intersect(dat_new2,dat_our)
num_uniq_new <- length(dat_new2) - length(overlap)

dat_plot <- list(A = 1:length(dat_new2), B = (num_uniq_new+1 ): (num_uniq_new + length(dat_our) ))  
names(dat_plot) <- c("aCM(new)","CM(ours)")
p[[1]] <- ggvenn(dat_plot, fill_alpha = 0) + 
  ggtitle("Atrial Cardiomyocyte \n VS \n Cardiomyocyte (all peaks)") + theme(plot.title = element_text(hjust = 0.5)) 


########
dat_new <- import("/project2/xinhe/xsun/atlas/peakdata_ren/celltype/Ventricular_cardiomyocyte.narrowPeak.bed", format = "BED",
                  extraCols = extraCols_narrowPeak)
dat_new2 <- unique(dat_new)

overlap <- GenomicRanges::intersect(dat_new2,dat_our)
num_uniq_new <- length(dat_new2) - length(overlap)

dat_plot <- list(A = 1:length(dat_new2), B = (num_uniq_new+1 ): (num_uniq_new + length(dat_our) ))  
names(dat_plot) <- c("vCM(new)","CM(ours)")
p[[2]] <- ggvenn(dat_plot, fill_alpha = 0) + 
  ggtitle("Ventricular Cardiomyocyte \n VS \n Cardiomyocyte (all peaks)") + theme(plot.title = element_text(hjust = 0.5)) 


########

ocr <- readRDS("/project2/gca/aselewa/heart_atlas_project/eQTL_enrich/OCR_disjoint_categories.grlist.rds")
dat_our <- ocr[[1]]
dat_new <- import("/project2/xinhe/xsun/atlas/peakdata_ren/specific/Atrial_cardiomyocyte.specific.elements.bed")
dat_new2 <- unique(dat_new)

overlap <- GenomicRanges::intersect(dat_new2,dat_our)
num_uniq_new <- length(dat_new2) - length(overlap)

dat_plot <- list(A = 1:length(dat_new2), B = (num_uniq_new+1 ): (num_uniq_new + length(dat_our) ))  
names(dat_plot) <- c("aCM(new)","CM(ours)")
p[[3]] <- ggvenn(dat_plot, fill_alpha = 0) + 
  ggtitle("Atrial Cardiomyocyte \n VS \n Cardiomyocyte (disjoint peaks)") + theme(plot.title = element_text(hjust = 0.5)) 


######
dat_new <- import("/project2/xinhe/xsun/atlas/peakdata_ren/specific/Ventricular_cardiomyocyte.specific.elements.bed")
dat_new2 <- unique(dat_new)

overlap <- GenomicRanges::intersect(dat_new2,dat_our)
num_uniq_new <- length(dat_new2) - length(overlap)

dat_plot <- list(A = 1:length(dat_new2), B = (num_uniq_new+1 ): (num_uniq_new + length(dat_our) ))  
names(dat_plot) <- c("vCM(new)","CM(ours)")
p[[4]] <- ggvenn(dat_plot, fill_alpha = 0) + 
  ggtitle("Ventricular Cardiomyocyte \n VS \n Cardiomyocyte (disjoint peaks)") + theme(plot.title = element_text(hjust = 0.5)) 


all <- grid.arrange(p[[1]],p[[3]],p[[2]],p[[4]], nrow = 2)

```

Our disjoint CM peaks are a subset of the CM peaks (all)

```{r echo=F, message=FALSE, warning=FALSE, paged.print=FALSE,fig.height=5, fig.width=5}
p<- list()

##########
ocr_all <- readRDS("/project2/gca/aselewa/heart_atlas_project/ArchR/ArchR_heart_latest_noAtrium/PeakCalls/DA_MARKERS_FDRP_1_log2FC_1.rds")
dat_our_all <- ocr_all[[1]]
ocr_disjoint <- readRDS("/project2/gca/aselewa/heart_atlas_project/eQTL_enrich/OCR_disjoint_categories.grlist.rds")
dat_our_disjoint  <- ocr_disjoint [[1]]

overlap <- findOverlaps(dat_our_all,dat_our_disjoint)
#overlap <- GenomicRanges::intersect(dat_our_all,dat_our_disjoint)
num_uniq_all <- length(dat_our_all) - length(overlap)

dat_plot <- list(A = 1:length(dat_our_all), B = (num_uniq_all+1 ): (num_uniq_all + length(dat_our_disjoint) ))  
names(dat_plot) <- c("CM_allpeak","CM_disjoint")
p <- ggvenn(dat_plot, fill_alpha = 0) + 
  ggtitle("Cardiomyocyte(all peaks) \n VS \n Cardiomyocyte (disjoint peaks)") + theme(plot.title = element_text(hjust = 0.5)) 
p
```





# Peak quality and overlap proportion (for all peaks)

We sorted our peak data by their quality (FDR). Then we computed the cumulative proportion that our peaks overlap with the new data. In the scatter plots below, each dot represents a peak range in our data set. Red dots mean the peak ranges do not overlap with the peaks in new data. The plot shows that the peaks have lower quality tend to not overlap with new data set since there are more dots in red when FDR getting larger. 

We only made plots for the pairs show correlation in heatmap.

```{r  message=FALSE, warning=FALSE, paged.print=FALSE,fig.height=10, fig.width=20}
ocr <- readRDS("/project2/gca/aselewa/heart_atlas_project/ArchR/ArchR_heart_latest_noAtrium/PeakCalls/DA_MARKERS_FDRP_1_log2FC_1.rds")
ocr_select <- c(1,1,2,3,5,6,7,8)
celltype_new <- c("Ventricular_cardiomyocyte","Atrial_cardiomyocyte","Endothelial","Fibroblast","Macrophage","Nervous","Smooth_muscle","Smooth_muscle")

folder_out <- "/project2/xinhe/xsun/atlas/cdf_plots/"

p <- list()
for (i in 1:length(ocr_select)) {
  
  dat_our <- ocr[[ocr_select[i]]]
  extraCols_narrowPeak <- c(signalValue = "numeric", pValue = "numeric",
                            qValue = "numeric", peak = "integer")
  dat_new <- import(paste0("/project2/xinhe/xsun/atlas/peakdata_ren/celltype/",celltype_new[i],".narrowPeak.bed"), format = "BED",
                    extraCols = extraCols_narrowPeak)
  dat_new <- unique(dat_new)
  
  df_our <- cbind(as.data.frame(dat_our@seqnames),as.data.frame(dat_our@ranges@start),as.data.frame(dat_our@ranges@width),dat_our$Log2FC,dat_our$FDR)
  colnames(df_our) <- c("chr","start","width","Log2FC","FDR")
  df_our$index <- 0
  
  overlap_index <- findOverlaps(dat_our,dat_new)
  index_our <- overlap_index@from
  index_our_uniq <- index_our[!duplicated(index_our)]
  df_our$index[index_our_uniq] <- 1
  
  df_our$persentage <- as.numeric(as.character(df_our$index))/sum(as.numeric(as.character(df_our$index)))
  df_our_sort <- df_our[order(as.numeric(as.character(df_our$FDR))),]
  df_our_sort$cumsum <- cumsum(df_our_sort$persentage)
  df_our_sort$order <- seq(1:nrow(df_our_sort))
  
  save(df_our_sort, file = paste0(folder_out,names(ocr)[ocr_select[i]],"_df.rdata")) 
  
  p[[i]] <- ggplot(df_our_sort, aes(x=order,y=cumsum)) + 
    theme_bw(base_line_size =0.3) +
    
    geom_point(aes(colour = factor(index))) +
    scale_color_manual("overlap",values=c("red", "grey"),labels = c("no", "yes")) +
    
    xlab("Order by FDR from our data (low to high)") + ylab("Cumulative percentage of overlapped peaks") +
    ggtitle(paste0(names(ocr)[ocr_select[i]],"(ours)","\n and \n ",celltype_new[i], "(new)")) +
    
    theme(plot.title = element_text(hjust = 0.5)) +
    
    annotate("text", x = length(dat_our) *0.2, y = 1, label = paste0("number of peaks_our = ", length(dat_our))) +
    annotate("text", x = length(dat_our) *0.2, y = 0.95, label = paste0("number of peaks_new = ", length(dat_new))) +
    annotate("text", x = length(dat_our) *0.2, y = 0.9, label = paste0("number of peaks_overlap = ", sum(df_our_sort$index)))
  
}

all <- grid.arrange(p[[1]],p[[2]],p[[3]],p[[4]],p[[5]],p[[6]],p[[7]],p[[8]], nrow = 2)

```

