---
title: "Label transfer (RNA-seq)"
author: "XSun"
date: "2022-08-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r, echo = FALSE}
suppressMessages(library(ComplexHeatmap))
```


# Introduction

We did label transfer for our study and [Litvinukova's study (Cells of the adult human heart)](https://www.nature.com/articles/s41586-020-2797-4).The RNA-seq data for the their study are available [here](https://www.heartcellatlas.org/#DataSources).

Overview of the two datasets:

The new data:

```{r echo=T, eval=FALSE, message=FALSE, warning=FALSE}
An object of class Seurat 
33538 features across 486134 samples within 1 assay 
Active assay: RNA (33538 features, 2000 variable features)
 2 dimensional reductions calculated: pca, umap
```

Our data

```{r echo=T, eval=FALSE, message=FALSE, warning=FALSE}
An object of class Seurat 
26537 features across 49359 samples within 1 assay 
Active assay: RNA (26537 features, 2000 variable features)
 3 dimensional reductions calculated: pca, harmony, umap
```

# Standard workflow of label transfer

We followed the [standard label transfer workflow](https://satijalab.org/seurat/archive/v3.0/integration.html) to do the label transfer. 

We have 49539 cells in total, 48359 of them were annotated same labels. 349 of the not matched cells fell into the "NotAssigned" category which is not included in our dataset. 

## Summary table

The figures in the table shows the number of cells. Rows are from our dataset, columns are from new dataset

```{r echo=F, message=FALSE, warning=FALSE}
summary<-read.table("/project2/xinhe/xsun/heart_atlas/2.label_transfer/data/summary.txt",header = T)
colnames(summary)[2] <- "vCardiomyocyte"
colnames(summary)[9] <- "Smooth Muscle"

rownames(summary) <- c("Cardiomyocyte", "Endothelial","Fibroblast","Lymphoid","Myeloid","Neuronal","Pericyte","Smooth Muscle")
DT::datatable(summary, options = list(pageLength =10))
```

## Heatmap

The heatmap shows the proportion of cells have the matched annotations. 

```{r echo=F, message=FALSE, warning=FALSE}
heat_map<-read.table("/project2/xinhe/xsun/heart_atlas/2.label_transfer/data/heatmap_matrix.txt",header = T)
#colnames(heat_map)[1] <- "Ventricular Cardiomyocyte"
colnames(heat_map)[1] <- "vCardiomyocyte"
colnames(heat_map)[8] <- "Smooth Muscle"

rownames(heat_map) <- c("Cardiomyocyte (8132)", "Endothelial (15336)","Fibroblast (11154)","Lymphoid (1002)","Myeloid (3356)","Neuronal (757)","Pericyte (7099)","Smooth Muscle (1376)")

Heatmap(matrix = heat_map, cluster_rows = F, 
        cluster_columns = F, show_row_names = T, row_names_side = "left",
        name = "Proportion \nof matched",
        col = circlize::colorRamp2(c(0, 0.1), c("white","firebrick")), ###modify the figures
        row_title = NULL,
        column_title = NULL,
        row_gap = unit(1, "mm"),
        column_gap = unit(1, "mm"),
        na_col = "white",
        use_raster = T)
```

